Do we need detailed demographic data to forecast plant population responses to climate change?

### 2015-01-12 manuscript/manuscript.Rmd
updated figure

Figures
-------
```{r figure_1, dependson="plot-options", fig.cap="Boxplots of model residuals for one-step-ahead forecasts at each observation year. Each one-step forecast was simulated `r nSims` times. Note that the y-axes vary across panels.", fig.width=5, fig.height=6, cache=FALSE, results='hide'}
resD <- readRDS("../montana/quadBM/simulations/quadBM_oneStep_Residuals.rds")
nSims <- length(unique(resD$sim))
diffD <- ddply(resD, c("year", "species"), summarise,
               avgDiff = mean(yearDiff))

ggplot(resD)+
  geom_hline(aes(yintercept=0))+
#   geom_line(data=diffD, aes(x=year, y=avgDiff, group=species), color="purple", alpha=0.5)+
#   stat_smooth(data=diffD, aes(x=year, y=avgDiff, group=species), color="lightblue", se = FALSE, size=1)+
  geom_boxplot(aes(x=year, y=covChangeResiduals), outlier.shape = 1, width=0.3, outlier.size = 0.5)+
  facet_grid(species~., scales = "free_y")+
  xlab("Year")+
  ylab("Residuals (% cover difference)")+
  theme_bw()+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

>>>>

Figures
-------
```{r figure_1, dependson="plot-options", fig.cap="Boxplots of model residuals for one-step-ahead forecasts at each observation year. Each one-step forecast was simulated `r nSims` times. Note that the y-axes vary across panels.", fig.width=8, fig.height=8, cache=FALSE, results='hide'}
resD <- readRDS("../montana/quadBM/simulations/quadBM_oneStep_ResidualsFullModel.rds")
resDny <- readRDS("../montana/quadBM/simulations/quadBM_oneStep_ResidualsNoYear.rds")
resD$simtype <- rep("full", nrow(resD))
resDny$simtype <- rep("noyear", nrow(resDny))
resDall <- rbind(resD, resDny)
# nSims <- length(unique(resD$sim))
diffD <- ddply(resDall, c("year", "species", "simtype"), summarise,
               avgDiff = mean(resids, na.rm = TRUE),
#                upDiff75 = quantile(resids, probs = 0.8775, names=FALSE, na.rm = TRUE),
#                dnDiff75 = quantile(resids, probs = 0.1225, names=FALSE, na.rm = TRUE),
               dnDiff95 = quantile(resids, probs = 0.025, names=FALSE, na.rm = TRUE),
               upDiff95 = quantile(resids, probs = 0.975, names=FALSE, na.rm = TRUE))
# ggplot(diffD, aes(color=simtype))+
#   geom_hline(aes(yintercept=0), linetype=2, color="grey50")+
#   geom_point(aes(x=year, y=avgDiff), size=3, position=position_dodge(width=0.5))+
#   geom_errorbar(aes(x=year, ymax=upDiff95, ymin=dnDiff95), width=0.00001, position=position_dodge(width=0.5))+
# #   geom_errorbar(aes(x=year, ymax=upDiff75, ymin=dnDiff75), size=1, width=0.00001, position=position_dodge(width=0.5))+
#   facet_grid(species~., scales = "free_y")+
#   xlab("Year")+
#   ylab("Residuals (% cover difference)")+
#   theme_bw()+
#   theme(axis.text.x = element_text(angle = 45, hjust = 1))
# 
# ggplot(resDall)+
#   geom_hline(aes(yintercept=0), linetype=2)+
# #   geom_line(data=diffD, aes(x=year, y=avgDiff, group=species), color="purple", alpha=0.5)+
# #   stat_smooth(data=diffD, aes(x=year, y=avgDiff, group=species), color="lightblue", se = FALSE, size=1)+
#   geom_boxplot(aes(x=year, y=resids, fill=simtype), outlier.shape = 1, width=0.3, outlier.size = 0.5)+
#   facet_grid(species~., scales = "free_y")+
#   xlab("Year")+
#   ylab("Residuals (% cover difference)")+
#   theme_bw()+
#   theme(axis.text.x = element_text(angle = 45, hjust = 1))
p1 <- ggplot(subset(resDall, species=="BOGR"))+
  geom_hline(aes(yintercept=0), linetype=2)+
  geom_boxplot(aes(x=year, y=resids, fill=simtype), outlier.shape = 1, outlier.size = 0, alpha=0.75, color="grey25")+
  xlab("Year")+
  ylab("Residuals (% cover difference)")+
  theme_bw()+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))+
  scale_fill_manual(values=c("darkgoldenrod1", "steelblue4"))+
  ggtitle(expression(italic("B. gracilis")))+
  guides(fill=FALSE)
p2 <- ggplot(subset(resDall, species=="HECO"))+
  geom_hline(aes(yintercept=0), linetype=2)+
  geom_boxplot(aes(x=year, y=resids, fill=simtype), outlier.shape = 1, outlier.size = 0, alpha=0.75, color="grey25")+
  xlab("Year")+
  ylab("Residuals (% cover difference)")+
  theme_bw()+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))+
  scale_fill_manual(values=c("darkgoldenrod1", "steelblue4"), labels=c("Year effects", "No year effects"))+
  ggtitle(expression(italic("H. comita")))+
  coord_cartesian(ylim = c(-25, 5))+
  guides(fill=guide_legend(title=""))+
  theme(legend.position=c(0.7,0.2))
p3 <- ggplot(subset(resDall, species=="PASM"))+
  geom_hline(aes(yintercept=0), linetype=2)+
  geom_boxplot(aes(x=year, y=resids, fill=simtype), outlier.shape = 1, outlier.size = 0, alpha=0.75, color="grey25")+
  xlab("Year")+
  ylab("Residuals (% cover difference)")+
  theme_bw()+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))+
  scale_fill_manual(values=c("darkgoldenrod1", "steelblue4"))+
  ggtitle(expression(italic("P. smithii")))+
  guides(fill=FALSE)+
  coord_cartesian(ylim = c(-2.5, 2.5))
p4 <- ggplot(subset(resDall, species=="POSE"))+
  geom_hline(aes(yintercept=0), linetype=2)+
  geom_boxplot(aes(x=year, y=resids, fill=simtype), outlier.shape = 1, outlier.size = 0, alpha=0.75, color="grey25")+
  xlab("Year")+
  ylab("Residuals (% cover difference)")+
  theme_bw()+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))+
  scale_fill_manual(values=c("darkgoldenrod1", "steelblue4"))+
  ggtitle(expression(italic("P. secunda")))+
  guides(fill=FALSE)+
  coord_cartesian(ylim = c(-50, 10))
pFinal <- grid.arrange(p1,p2,p3,p4, ncol=2, nrow=2)
pFinal

### 2015-01-12 manuscript/manuscript.tex
updated figure

Aggregate measures of individual plant performance, such as those
typically collected as part of large-scale census efforts, offer an
alternative to detailed demographic data for modeling populations (Clark
and Bjørnstad 2004, Freckleton et al. 2011). Such population-level data
will never match the precision of individual-level data, but it is more
feasible to attain a broad coverage sample when collecting coarse-scale
data. This presents a difficult trade-off: on the one hand,
individual-level data leads to more reliable models; on the other hand,
population-level data leads to models that will produce less precise
predicitons but can be applied over greater spatial and temporal
extents. An open question is how well models based on population-level
data compare to models based on individual-level data.

To date, relatively few studies have tried to model populations based on
data other than detailed individual-level data. An important exception
is an effort by Taylor and Hastings (2004) to model the population
growth rate of an invasive species to investigate the best strategies
for invasion control. They used a ``density-structured'' model where the
state variable is a discrete density state rather than a continuous
density measure. Building on this work, Freckleton et al. (2011) showed
that density-structured models compare well to continuous models in
theory, and Queenborough et al. (2011) showed the application of such
methods in a study on arable weeds. In particular, Queenborough et al.
(2011) provide emprical evidence that density-structured models are
capable of reproducing population dynamics, even if some precision is
lost when compared to fully continuous models. Thus, population models
based on coarse, population-level data show promise for producing
ecological forecasts at landscape and regional scales (Queenborough et
al. 2011). However, none of these models included environmental
covariates.

@@

Our demographic data comes from the Fort Keogh Livestock and Range
Research Laboratory in eastern Montana's northern mixed prairie near
Miles City, Montana, USA ($46^{\circ}$ 19' N, $105^{\circ}$ 48' W). The
dataset is freely available on Ecological Archives (Anderson et al.
2011), and interested readers should refer to the metadata therin for a
complete description. The site is about 800 m above sea level and mean
annual precipitation (1878-2009) is 334 mm, with most annual
precipitaion falling from April through September. The site is grass
dominated and, for the purposes of our study, we focus on the four most
abundant graminoid species: \emph{Bouteloua gracilis} (BOGR),
\emph{Hesperostipa comata} (HECO), \emph{Pascopyrum smithii} (PASM), and
\emph{Poa secunda} (POSE).

From 1932 to 1945 individual plants were identified and mapped annualy
in 44 1-$\text{m}^2$ quadrats using a pantograph. The quadrats were
distributed in six pastures, each assigned a grazing treatment of light
(1.24 ha/animal unit month), moderate (0.92 ha/aum), and heavy (0.76
ha/aum) stocking rates (two pastures per treatment). In this analysis we
account for potential differences among the grazing treatments, but do
not focus on grazing$\times$climate interactions. The annual maps of the
quadrats were digitized and the fates of individual plants tracked and
extracted using a computer program. Daily climate data, which we
aggregated into climate variables of interest, are available for the
duration of the data collection period (1932 - 1945) from the Miles City
airport, Wiley Field, 9 km from the study site.

@@

At both levels of inference (individual and quadrat), the building
blocks of our population models are vital rate regressions. For
individual level data we fit models for survival, growth, and
recruitment of new individuals for each species. At the quadrat level we
fit analagous models of extinction probability, percent cover
increase/decrease, and quadrat colonization for each species. We
describe the statistical models seprately since fitting the models
required different approaches at the individual and quadrat levels. All
models contain four climate covariate that we chose \emph{a priori}:
fall through spring precipitation at \emph{t}-1 and \emph{t}-2 (ppt1 and
ppt2, respectively) and mean spring temperature at \emph{t}-1 and
\emph{t}-2 (TmeanSpr1 and TmeanSpr2, respectively), where \emph{t} is
the observation year.

@@

We retain the abbreviations $S$, $G$, and $R$ from the analagous
processes at the individual level. The vital rate models at the quadrat
level are all based on quadrat proportional cover. Also, at the quadrat
level we do not need to explicitly include a density dependent term.
Since we are modeling proportional cover, we essentially get
density-dependence for ``free'' when proportional cover is included as a
covariate. That is, density-dependence emerges because quadrats with low
cover gerenally increase in cover whereas quadrats with high cover
generally decrease in cover.

@@

Our Bayesian approach to fitting the vital rate models required choosing
appropriate priors for unknown parameters and deciding which, if any, of
those prior should be hierarchical. We decided to fit models where all
terms except climate covariates were fit by species, while the climate
covariates were fit heierarchically where species-specific coefficients
were drawn from a shared `global' coefficient distribution. We did so
for two reasons: (1) the four focal species are all perrennial grasses
that we expect to respond similarly to climate covariates, and (2)
convergence of climate effects at the quadrat level was much easier to
achieve when we modeled these terms hierarchically, allowing them to
``share'' statistical strenght via partial pooling (Gelman and Hill
2007). So, climate effects were modeled as:

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\itemsep1pt\parskip0pt\parsep0pt
\item
  Climate detections
\item
  Limitations:
\end{enumerate}

>>>>

Aggregate measures of individual plant performance, such as those
typically collected as part of large-scale census efforts, offer an
alternative to detailed demographic data for modeling populations (Clark
and Bjørnstad 2004, Freckleton et al. 2011). Such population-level data
will never match the precision of individual-level data, but it is more
feasible to attain a broad coverage sample when collecting coarse-scale
data. This presents a difficult trade-off: on the one hand,
individual-level data leads to more reliable models; on the other hand,
population-level data leads to models that will produce less precise
predictions but can be applied over greater spatial and temporal
extents. An open question is how well models based on population-level
data compare to models based on individual-level data.

To date, relatively few studies have tried to model populations based on
data other than detailed individual-level data. An important exception
is an effort by Taylor and Hastings (2004) to model the population
growth rate of an invasive species to investigate the best strategies
for invasion control. They used a ``density-structured'' model where the
state variable is a discrete density state rather than a continuous
density measure. Building on this work, Freckleton et al. (2011) showed
that density-structured models compare well to continuous models in
theory, and Queenborough et al. (2011) showed the application of such
methods in a study on arable weeds. In particular, Queenborough et al.
(2011) provide empirical evidence that density-structured models are
capable of reproducing population dynamics, even if some precision is
lost when compared to fully continuous models. Thus, population models
based on coarse, population-level data show promise for producing
ecological forecasts at landscape and regional scales (Queenborough et
al. 2011). However, none of these models included environmental
covariates.

@@

Our demographic data comes from the Fort Keogh Livestock and Range
Research Laboratory in eastern Montana's northern mixed prairie near
Miles City, Montana, USA ($46^{\circ}$ 19' N, $105^{\circ}$ 48' W). The
dataset is freely available on Ecological Archives (Anderson et al.
2011), and interested readers should refer to the metadata therein for a
complete description. The site is about 800 m above sea level and mean
annual precipitation (1878-2009) is 334 mm, with most annual
precipitation falling from April through September. The site is grass
dominated and, for the purposes of our study, we focus on the four most
abundant graminoid species: \emph{Bouteloua gracilis} (BOGR),
\emph{Hesperostipa comata} (HECO), \emph{Pascopyrum smithii} (PASM), and
\emph{Poa secunda} (POSE).

From 1932 to 1945 individual plants were identified and mapped annually
in 44 1-$\text{m}^2$ quadrats using a pantograph. The quadrats were
distributed in six pastures, each assigned a grazing treatment of light
(1.24 ha/animal unit month), moderate (0.92 ha/aum), and heavy (0.76
ha/aum) stocking rates (two pastures per treatment). In this analysis we
account for potential differences among the grazing treatments, but do
not focus on grazing$\times$climate interactions. The annual maps of the
quadrats were digitized and the fates of individual plants tracked and
extracted using a computer program. Daily climate data, which we
aggregated into climate variables of interest, are available for the
duration of the data collection period (1932 - 1945) from the Miles City
airport, Wiley Field, 9 km from the study site.

@@

At both levels of inference (individual and quadrat), the building
blocks of our population models are vital rate regressions. For
individual level data we fit models for survival, growth, and
recruitment of new individuals for each species. At the quadrat level we
fit analogous models of extinction probability, percent cover
increase/decrease, and quadrat colonization for each species. We
describe the statistical models separately since fitting the models
required different approaches at the individual and quadrat levels. All
models contain four climate covariate that we chose \emph{a priori}:
fall through spring precipitation at \emph{t}-1 and \emph{t}-2 (ppt1 and
ppt2, respectively) and mean spring temperature at \emph{t}-1 and
\emph{t}-2 (TmeanSpr1 and TmeanSpr2, respectively), where \emph{t} is
the observation year.

@@

We retain the abbreviations $S$, $G$, and $R$ from the analogous
processes at the individual level. The vital rate models at the quadrat
level are all based on quadrat proportional cover. Also, at the quadrat
level we do not need to explicitly include a density dependent term.
Since we are modeling proportional cover, we essentially get
density-dependence for ``free'' when proportional cover is included as a
covariate. That is, density-dependence emerges because quadrats with low
cover generally increase in cover whereas quadrats with high cover
generally decrease in cover.

@@

Our Bayesian approach to fitting the vital rate models required choosing
appropriate priors for unknown parameters and deciding which, if any, of
those prior should be hierarchical. We decided to fit models where all
terms except climate covariates were fit by species, while the climate
covariates were fit hierarchically where species-specific coefficients
were drawn from a shared `global' coefficient distribution. We did so
for two reasons: (1) the four focal species are all perennial grasses
that we expect to respond similarly to climate covariates, and (2)
convergence of climate effects at the quadrat level was much easier to
achieve when we modeled these terms hierarchically, allowing them to
``share'' statistical strength via partial pooling (Gelman and Hill
2007). So, climate effects were modeled as:

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\itemsep1pt\parskip0pt\parsep0pt
\item
  Climate detection
\item
  Limitations:
\end{enumerate}
